Approximation as Differentiation

### 2025-07-08 main/models.tex
[ tex ] talk about chain rule more
[ tex ] refine the discussion of Fam(C) a bit

\section{Constructing Models of \GPS for a Total Language}
\label{sec:models-of-total-gps}

The previous section concluded that L-posets and stable functions give
a model of \GPS analogous to manifolds and smooth functions. However,
we noted a conceptual shortcoming of this model, for the purposes of
modelling total computations, that proper values and their
approximations live in the same category. In this section, we propose
a model for total \GPS based on the \emph{Category of Families}
construction. This construction, adn the more general Grothendieck
construction, has been previously used by V{\'a}k{\'a}r and
collaborators \bob{CITE} to model automatic differentiation for
higher-order programs on the reals. We rely on some of their results,
and discuss the commonalities as we go.

@@

As we explained in \secref{stab}, L-posets can be seen as partially
ordered sets where every principal downset $\downset{x}$ is a bounded
lattice. That is, for an element $x$, $\downset{x}$ is its lattice of
approximations/tangents. The shortcoming of this setup is that proper
values and their approximations live in the same set. We fix this by
changing our model to one where we have sets $X$ of values, and for
each $x \in X$, a bounded lattice $\partial X(x)$ of approximations of
$x$. This construction is an instance of the general \emph{Category of
  Families} construction:

\begin{definition}
  Let $\cat{C}$ be a category. The \emph{Category of Families} over
  $\cat{C}$, $\Fam(\cat{C})$, has as objects pairs $(X, \partial X)$,
  where $X$ is a set and $\partial X : X \to \cat{C}$ is an
  $X$-indexed family of objects in $\cat{C}$. A morphism
  $f : (X, \partial X) \to (Y, \partial Y)$ consists of a pair of a
  function $f : X \to Y$ and a familiy of morphisms of $\cat{C}$,
  $\partial f : (x : X) \to \cat{C}(\partial X(x), \partial Y(f\,x))$.
\end{definition}

Composition $f \circ g $ of morphisms
$f : (Y, \partial Y) \to (Z, \partial Z)$ and
$g : (X, \partial X) \to (Y, \partial Y)$ in this category is given by
normal function composition on the set components, and
$\partial (f \circ g)(x) = \partial f(f, x) \circ \partial g(x)$,
where the latter composition is in $\cat{C}$. The fact that
composition of morphisms in $\Fam(\cat{C})$ resembles the chain rule
from calculus is essential to its use in modelling automatic
differentiation. Indeed, if we let $\FinVect$ be the category of
finite dimensional real vector spaces, then:

@@

For \GPS we replace the category $\FinVect$ with $\LatGal$:

\begin{definition}
  The category $\LatGal$ has bounded lattices as objects and Galois
  connections as morphisms, with the right adjunct going in the
  ``forward'' direction.
\end{definition}

@@

\begin{proposition}
  There is a faithful functor $\Lposet \to \Fam(\LatGal)$ that maps an
  L-poset $X$ to $(X, \lambda x. \downset{x})$ and stable functions
  $f$ to $(f, \lambda x. (f_x, f^*_x))$.
\end{proposition}

Despite its similarity, this proposition has a lesser status than
\propref{embed-manifolds} because it is not clear that the category
$\Lposet$ is in any way a canonical definition of approximable sets
and functions with approximation derivatives. We could restrict
$\Lposet$ to morphisms that preserve maximal elements and take
L-posets $X$ to $(\mathrm{Max}(M), \lambda x. \downset{x})$. However,
our current hypothesis is that $\Fam(\LatGal)$, where values and their
approximations are separated by construction is the primary model of
semantic \GPS. We now investigate the categorical properties of this
category, with a view to modelling a higher-order total programming
language.

@@

\begin{proposition}
  For any $\cat{C}$, $\Fam(\cat{C}$ has all coproducts, which can be
  given on objects by:
  \begin{displaymath}
    \coprod_i (X_i, \partial X_i) = (\coprod_i X_i, \lambda (i, x_i).\, \partial X_i(x))
  \end{displaymath}
  Coproducts in $\Fam(\cat{C})$ are \emph{extensive}
  \cite{carboni-lack-walters93} (Proposition 2.4).
\end{proposition}

@@

\begin{proposition}
  If $\cat{C}$ has finite products, then so does $\Fam(\cat{C})$. On
  objects, binary products can be defined by:
  \begin{displaymath}
    (X, \partial X) \times (Y, \partial Y) = (X \times Y, \lambda (x, y). \partial X(x) \times \partial Y(y))
  \end{displaymath}
  Since $\Fam(\cat{C})$ is extensive, products and coproducts always
  distribute.
\end{proposition}

@@

In the case of $\FinVect$, the solution is to expand to the category
of all vector spaces $\Vect$, where infinite direct products
exist. Note that these infinite products are not biproducts because
the vector space operations themselves are finitary. This is the
solution that \cite{vakar2021} uses for the semantics of forward
($\Fam(\Vect)$) and reverse ($\Fam(\Vect^\op)$) automatic
differentiation for higher order programs. Since the forward and
reverse derivatives of a smooth map intrinsically defined,
\citet{vakar2021}'s correctness theorem can show that

For $\LatGal$ we could expand to the category of complete lattices and
Galois connections between them. From a classical mathematical point
of view, this would give a model of \GPS that would likely be suitable
for reasoning programs' behaviour and their forward and backward
approximations. However, in terms of building an executable model
inside the Agda proof assistant, and with an eye toward implementation
strategies, we seek a more down-to-earth (i.e., predicative and
constructive) solution. We will do this by separating the forward and
backward parts of the Galois connections to act independently by
moving to the product category $\MeetSLat \times
\JoinSLat^\op$. Objects in this category consist of \emph{separate}
meet- and join-semi-lattices and potentially unrelated forward
meet-preserving and backward join-preserving maps. We first check that
this category satisfies the hypotheses of \thmref{fam-closed}:

>>>>

\section{Models of \GPS for a Total Language}
\label{sec:models-of-total-gps}

The previous section concluded that L-posets and stable functions give
a model of \GPS analogous to manifolds and smooth functions. However,
we noted a conceptual shortcoming of this model, for the purposes of
modelling total computations, that proper values and their
approximations live in the same category. In this section, we propose
a model for total \GPS based on the \emph{Category of Families}
construction. This construction, and the more general Grothendieck
construction, has been previously used by V{\'a}k{\'a}r and
collaborators \bob{CITE} to model automatic differentiation for
higher-order programs on the reals. We reuse some of their results,
and discuss the commonalities as we go.

@@

L-posets are partially ordered sets where every principal downset
$\downset{x}$ is a bounded lattice of approximations/tangents. As we
explained in \secref{stab}, the shortcoming of this setup is that
proper values and their approximations live in the same set. We fix
this by changing our model to one where we have sets $X$ of values,
and for each $x \in X$, a bounded lattice $\partial X(x)$ of
approximations of $x$. This construction is an instance of the general
\emph{Category of Families} construction:

\begin{definition}
  Let $\cat{C}$ be a category. The \emph{Category of Families} over
  $\cat{C}$, $\Fam(\cat{C})$, has as objects pairs $(X, \partial X)$,
  where $X$ is a set and $\partial X : X \to \cat{C}$ is an
  $X$-indexed family of objects in $\cat{C}$. A morphism
  $f : (X, \partial X) \to (Y, \partial Y)$ consists of a pair of a
  function $f : X \to Y$ and a familiy of morphisms of $\cat{C}$,
  $\partial f : \Pi_{x : X}.\,\cat{C}(\partial X(x), \partial Y(f\,x))$.
\end{definition}

The reason for choosing the $\Fam$ construction is that composition in
this category is an abstract version of the chain rule that we have
seen in \remref{chain-rule}, \remref{chain-rule-cm}, and
\remref{chain-rule-stable}. Composition $f \circ g $ of morphisms
$f : (Y, \partial Y) \to (Z, \partial Z)$ and
$g : (X, \partial X) \to (Y, \partial Y)$ in this category is given by
normal function composition on the set components, and
$\partial (f \circ g)(x) = \partial f(f, x) \circ \partial g(x)$,
where the latter composition is in $\cat{C}$.

The fact that morphisms in $\Fam(C)$ compose according to a chain rule
means that the categories we have seen so far embed into $\Fam(C)$ for
appropriate $C$. If we let $\FinVect$ be the category of finite
dimensional real vector spaces, then:

@@

For the categories $\CM$ and $\Lposet$, we pick the appropriate
categories of partial orders and monotone maps:

\begin{definition}
  \begin{enumerate}
  \item The category $\LatGal$ has bounded lattices as objects and
    Galois connections as morphisms, with the right adjunct going in
    the ``forward'' direction.
  \item The category $\MeetSLat$ has meet-semi-lattices with top as
    objects and monotone finite meet preserving functions as
    morphisms.
  \item The category $\JoinSLat$ has join-semi-lattices with bottom as
    objects and monotone finite join preserving functions as
    morphisms.
  \end{enumerate}
\end{definition}

@@

\begin{proposition}
  \label{prop:embed-stable}
  There is a faithful functor $\Lposet \to \Fam(\LatGal)$ that maps an
  L-poset $X$ to $(X, \lambda x. \downset{x})$ and stable functions
  $f$ to $(f, \lambda x. (f_x, f^*_x))$. Likewise, there is a faithful
  functor $\CM \to \Fam(\MeetSLat)$.
\end{proposition}

Despite its similarity, this proposition has a lesser status than
\propref{embed-manifolds} because it is not clear that the category
$\Lposet$ (or $\CM$) is a canonical definition of approximable sets
and functions with approximation derivatives, as we discussed at the
end of \secref{stab}.
% We could restrict $\Lposet$ to morphisms that preserve maximal
% elements and take L-posets $X$ to
% $(\mathrm{Max}(M), \lambda x. \downset{x})$.
Our current hypothesis is that $\Fam(\LatGal)$, where values and their
approximations are separated by construction, is the natural model of
semantic \GPS in a total setting. We now investigate some categorical
properties of this category, with a view to modelling a higher-order
total programming language in \secref{language}.

@@

\begin{proposition}
  For any $\cat{C}$, $\Fam(\cat{C})$ has all coproducts, which can be
  given on objects by:
  \begin{displaymath}
    \coprod_i (X_i, \partial X_i) = (\coprod_i X_i, \lambda (i, x_i).\, \partial X_i(x))
  \end{displaymath}
  Coproducts in $\Fam(\cat{C})$ are \emph{extensive}
  \cite{carboni-lack-walters93} (Proposition 2.4).
\end{proposition}

@@

\begin{proposition}
  If $\cat{C}$ has finite products, then so does $\Fam(\cat{C})$. On
  objects, binary products can be defined by:
  \begin{displaymath}
    (X, \partial X) \times (Y, \partial Y) = (X \times Y, \lambda (x, y). \partial X(x) \times \partial Y(y))
  \end{displaymath}
  Since $\Fam(\cat{C})$ is extensive, products and coproducts
  distribute.
\end{proposition}

@@

In the case of $\FinVect$, the solution is to expand to the category
of all vector spaces $\Vect$, where infinite direct products
exist. Note that these infinite products are not biproducts because
the vector space operations themselves are finitary. This is the
solution that \cite{vakar2021} uses for the semantics of forward
($\Fam(\Vect)$) and reverse ($\Fam(\Vect^\op)$) automatic
differentiation for higher order programs. Since the forward and
reverse derivatives of a smooth map intrinsically defined,
\citet{vakar2021}'s correctness theorem shows that, for programs with
first-order type, the interpretation in $\Fam(\Vect)$ correctly yields
the forward derivative of the defined function on the reals (and
reverse derivative for $\Fam(\Vect^\op)$.

For $\LatGal$ we could expand to the category of complete lattices and
Galois connections between them. From a classical mathematical point
of view, this would give a model of \GPS that would be suitable for
reasoning about programs' behaviour and their forward and backward
approximations. However, in terms of building an executable model
inside the Agda proof assistant, and with an eye toward implementation
strategies, we seek a finitary solution. (Note that the solution of
moving to complete lattices is very different to moving to arbitrary
dimension vector spaces: in the former we have infinitary operations,
while the latter still has only finitary operations.)

We will avoid the need for infinitary operations by separating the
forward and backward parts of the Galois connections to act
independently by moving to the product category
$\MeetSLat \times \JoinSLat^\op$. Objects in this category consist of
\emph{separate} meet- and join-semi-lattices and potentially unrelated
forward meet-preserving and backward join-preserving maps. We first
check that this category satisfies the hypotheses of
\thmref{fam-closed}:
